Enforcing Boundaries: China's Overseas NGO Law and Operational Constraints for Global Civil Society

### 2024-02-24 manuscript/manuscript.qmd
Expanded introduction to E1

---
title: "Enforcing Boundaries: China's Overseas NGO Law and Operational Constraints for Global Civil Society"
short-title: "Enforcing Boundaries"
date: "2024-02-23"
author:
- name: Meng Ye
  affiliation: Georgia State University
  email: mye2@gsu.edu
  orcid: 0000-0002-6552-8338
- name: Andrew Heiss
  affiliations: 
    - id: gsu
      name: Georgia State University
      department: Andrew Young School of Policy Studies, Department of Public Management and Policy
      address: "55 Park Place NE, #464"
      city: Atlanta
      region: GA
      country: USA
      postal-code: 30303
  email: aheiss@gsu.edu
  orcid: 0000-0002-3948-3914
  attributes:
    corresponding: true
correspondence-prefix: "Correspondence concerning this article should be addressed to"
abstract: |
  China's 2016 Overseas NGO (ONGO) Law is part of a larger global trend of increased legal restrictions on international nongovernmental organizations (INGOs). A growing body of research analyzes the broad effects of this crackdown on INGOs, finding a divergence in formal *de jure* laws and the *de facto* implementation of those laws. The causes and mechanisms of this divergence remain less explored. Why do authoritarian governments allow—and often collaborate—with some INGOs while harshly regulating or expelling others? What determines the openness of the practical legal operating environment for INGOs? In this paper, we use the case of China to explore how political demands to both restrict and embrace INGOs have shaped the international nonprofit sector in the five years since the ONGO Law came into effect. We argue that in an effort to bolster regime stability, governments use civil society laws as policy tools to influence INGO behavior. We find that INGO issue areas, missions, and pre-existing relationships with local government officials influence the degree of operating space available for INGOs. We test this argument with a mixed methods research design, combining Bayesian analysis of administrative data from all formally registered INGOs with a comparative case study of two environmental INGOs. Our findings offer insights into the practical effects of INGO restrictions and the dynamics of closing civic space worldwide.
additional-info: |
  **Compliance with Ethical Standards:** This study does not involve human subjects, so the need for participant protection measures such as informed consent, confidentiality, and privacy safeguards was not applicable.

@@

We sketch this relationship in @fig-theory, which demonstrates several key determinants of INGO operating space in authoritarian contexts. The left half of the diagram illustrates how the interplay of government preferences and INGO issue areas shapes how the government enforces its civil society laws, which in turn helps determine how much operating space an INGO can enjoy. In general, the alignment of an INGO's programming with government preferences determines how laws are enforced, which in turn helps determine how much operating space an INGO can enjoy. INGO laws that are intentionally ambiguous with large room for discretion can serve as policy tools for authoritarian states to seek out INGOs that provide direct benefits, while avoiding INGOs that threaten their stability [@LiFarid:2023; @ChaudhryHeiss:2021]. For instance, states in need of foreign aid and development assistance will be more open to INGOs working on those issues, while states looking for increased trade will seek out INGOs focused on economic partnerships. Because their primary issue areas are aligned with regime preferences, the practical *de facto* enforcement of civil society will be less strict and INGOs will benefit from greater operational space and face fewer restrictions. In contrast, states engaged in human rights abuses or fearful of foreign influence over other domestic policies will reduce---or even close---the operational space allowed for INGOs addressing these issues. Notably, the arrow between government preferences and issue area is bidirectional. Restrictive civil society regulations create a population of allowed INGOs aligned with government preferences, either by attracting safe and noncontentious organizations, or by encouraging potentially contentious organizations to "tame" their programming [@Bush:2015]. Conversely, once allowed in the country, INGOs can reshape government views and policies regarding both their given issue areas and civil society regulations more broadly [@Teets:2014; @Heiss:2019a].

The right half of @fig-theory illustrates alternative determinants of INGO operational space that lie outside of organizations' potential contentiousness or relationships with the government. INGO preferences—that is, the strategy, mission, and goals—influence how much operational space is available in host countries. For instance, many INGOs have missions focused on local issues and communities—they work on specific projects with naturally limited scopes and have no need to expand their work. An INGO's available resources also determine its in-country flexibility. An organization with more funding, more employees, more connections with local regulators or politicians, or other types of resources can benefit from additional flexibility when working in its host country by, for instance, enjoying greater geographic access, more leeway in adherence to regulations, or a wider menu of possible strategies available to accomplish its mission. Organizations with fewer resources, on the other hand, will have more limited operational space.

To test the relationships between government preferences, INGO issue area, and INGO operational space, we apply this conceptual model to China's ONGO law. Previous research has looked at *A posteriori* (notes: I don't know if it is a correct use of latin words, I want to say how the INGO sector is like in a timeframe after the law) effects of civil society laws on a global scale [@LianMurdie:2023; @Chaudhry:2022], which naturally blurs the specific policy mechanisms regimes use for controlling and leveraging global civil society. By analyzing the population of formally registered INGOs between 2017–2021, we can more precisely explore how a single formal restrictive INGO law has shaped the operational space for foreign organizations working in the country. Specifically, we can observe exactly how the Chinese government has used the RO registration process to strategically limit the legal presence and potential reach of INGOs and reshape their issue areas. We conducted both a descriptive analysis of the composition of the issue areas of INGOs successfully registered under the INGO law and a regression analysis of the relationship between INGO issue areas and their operational space.

We use the number of provinces an INGO is authorized to work in as a proxy for organizational flexibility. The rationale is it is preferable for INGOs to have a larger approved geographical operation scope on paper though they may not actually conduct activities there. Case studies showed that INGOs such as Oxfam and World Vision went through the cumbersome process to register multiple ROs to cover all the provinces they need to work in [@Li:2020]. Their ideal goal is registering an office with authorization to work across China Mainland but was not approved so, ie. encountered more restriction with each registration. Moreover, having a larger number approved scope is important for INGOs that want to be prepared to changing target community needs in emergency situations as the COVID-19 and are prudent in legal risk management when the definition of activities is ambiguous under the INGO law.

Thus, we propose a set of expectations based on the conceptual model in @fig-theory and rationales explained. First, we expect that the *de facto* operational space of INGOs varies by INGO issue areas:

> **E~1~**: Because issue areas reflect varied level of contentiousness and government preferences, the number of provinces INGOs are approved to operate in is associated with their issue areas.

@@

Before testing these expectations, we must first note a few important caveats. As discussed previously, international NGOs have two avenues for legally working in China: (1) formal registration as an RO and (2) ad hoc temporary activity (TA) status. The RO process is more costly and requires far more administrative and legal work, and as a result, nearly twice as many organizations have used temporary activities rather than formal registration. In this study, however, we do not look at patterns in TA use to explore INGO operational space. There is an incredible amount of variation in how TAs are used. Some organizations submit dozens of TA requests, possibly as a type of workaround to obtaining RO status, while others use them as true temporary activities, submitting requests for one-off conferences or workshops. INGOs using TAs thus do not work well as a test case of the relationship between INGO operational space and the de facto implementation of the ONGO law. More permanent ROs are more appropriate. Additionally, we cannot observe or measure the activities of INGOs that have not attempted to establish operations in the country or who left the country following the implementation of the ONGO law in 2017. The population of INGOs working with full, permanent legal status is thus limited only to organizations that have already been vetted as "safe" and acceptable by the government. Even under these conditions, though, we should expect to see variation in operational space, due to heterogeneity in both (1) government preferences and INGO issues (i.e. not all registered INGOs work on the same issues, and not all issues are of equal value or importance to the regime), and (2) INGO preferences and resources (i.e. not all INGOs have the same goals, missions, or resources).

@@

*add the discussion of "significance level, or the interpretation of credible intervals. That is move some of the notes to the figure to the main text?"*

>>>>

---
title: "Enforcing Boundaries: China's Overseas NGO Law and Operational Constraints for Global Civil Society"
short-title: "Enforcing Boundaries"
date: "2024-02-23"
author:
- name: Meng Ye
  affiliation: Georgia State University
  email: mye2@gsu.edu
  orcid: 0000-0002-6552-8338
- name: Andrew Heiss
  affiliations: 
    - id: gsu
      name: Georgia State University
      department: Andrew Young School of Policy Studies, Department of Public Management and Policy
      address: "55 Park Place NE, #464"
      city: Atlanta
      region: GA
      country: USA
      postal-code: 30303
  email: aheiss@gsu.edu
  orcid: 0000-0002-3948-3914
  attributes:
    corresponding: true
correspondence-prefix: "Correspondence concerning this article should be addressed to"
abstract: |
  China's 2016 Overseas NGO (ONGO) Law is part of a larger global trend of increased legal restrictions on international nongovernmental organizations (INGOs). A growing body of research analyzes the broad effects of this crackdown on INGOs, finding a divergence in formal *de jure* laws and the *de facto* implementation of those laws. The causes and mechanisms of this divergence remain less explored. Why do authoritarian governments allow---and often collaborate---with some INGOs while harshly regulating or expelling others? What determines the openness of the practical legal operating environment for INGOs? In this paper, we use the case of China to explore how political demands to both restrict and embrace INGOs have shaped the international nonprofit sector in the five years since the ONGO Law came into effect. We argue that in an effort to bolster regime stability, governments use civil society laws as policy tools to influence INGO behavior. We find that INGO issue areas, missions, and pre-existing relationships with local government officials influence the degree of operating space available for INGOs. We test this argument with a mixed methods research design, combining Bayesian analysis of administrative data from all formally registered INGOs with a comparative case study of two environmental INGOs. Our findings offer insights into the practical effects of INGO restrictions and the dynamics of closing civic space worldwide.
additional-info: |
  **Compliance with Ethical Standards:** This study does not involve human subjects, so the need for participant protection measures such as informed consent, confidentiality, and privacy safeguards was not applicable.

@@

We sketch this relationship in @fig-theory, which demonstrates several key determinants of INGO operating space in authoritarian contexts. The left half of the diagram illustrates how the interplay of government preferences and INGO issue areas shapes how the government enforces its civil society laws, which in turn helps determine how much operating space an INGO can enjoy. In general, the alignment of an INGO's programming with government preferences determines how laws are enforced, which in turn helps determine how much operating space an INGO can enjoy. INGO laws that are intentionally ambiguous with large room for discretion can serve as policy tools for authoritarian states to seek out INGOs that provide direct benefits, while avoiding INGOs that threaten their stability [@LiFarid:2023; @ChaudhryHeiss:2021]. For instance, states in need of foreign aid and development assistance will be more open to INGOs working on those issues, while states looking for increased trade will seek out INGOs focused on economic partnerships. Because their primary issue areas are aligned with regime preferences, the practical *de facto* enforcement of civil society will be less strict and INGOs will benefit from greater operational space and face fewer restrictions. In contrast, states engaged in human rights abuses or fearful of foreign influence over other domestic policies will reduce---or even close---the operational space allowed for INGOs addressing these issues. Notably, the arrow between government preferences and issue area is bidirectional. Restrictive civil society regulations create a population of allowed INGOs aligned with government preferences, either by attracting safe and noncontentious organizations, or by encouraging potentially contentious organizations to "tame" their programming [@Bush:2015]. Conversely, once allowed in the country, INGOs are not always passive victims of regulations and can, under particular conditions, reshape government views and policies regarding both their given issue areas and civil society regulations more broadly [@Teets:2014; @Heiss:2019a].

The right half of @fig-theory illustrates alternative determinants of INGO operational space that lie outside of organizations' potential contentiousness or relationships with the government. An INGO's preferences---that is, its strategy, mission, and goals---also influence how much operational space is available in host countries. For instance, many INGOs have missions focused on local issues and communities---they work on specific projects with naturally limited scopes and have no need to expand their work. The nature of an organization's programming also influences its reach and flexibility: a human rights advocacy organization interested in holding a national government accountable for civil liberties abuses will have a much broader geographic and political scope---and thus require greater operational flexibility---than an education organization interested in supporting a handful of primary schools. An INGO's available resources also determine its in-country flexibility. An organization with more funding, more employees, more connections with local regulators or politicians, or other types of resources can benefit from additional flexibility when working in its host country by, for instance, enjoying greater geographic access, more leeway in adherence to regulations, or a wider menu of possible strategies available to accomplish its mission. Organizations with fewer resources, on the other hand, will have more limited operational space. These organization-level determinants can be seen as independent of government preferences

To test the relationships between government preferences, INGO issue area, INGO characteristics, and operational space, we apply this conceptual model to China's 2017 ONGO law. Previous research has looked at *A posteriori* (notes: I don't know if it is a correct use of latin words, I want to say how the INGO sector is like in a timeframe after the law) effects of civil society laws on a global scale [@LianMurdie:2023; @Chaudhry:2022], which naturally blurs the specific policy mechanisms regimes use for controlling and leveraging global civil society. By analyzing the population of formally registered INGOs between 2017–2021, we can more precisely explore how a single formal restrictive INGO law has shaped the operational space for foreign organizations working in the country. Specifically, we can observe exactly how the Chinese government has used the RO registration process to strategically limit and reshape these organizations' operational space. We conducted both a descriptive analysis of the composition of the issue areas of INGOs successfully registered under the INGO law and a regression analysis of the relationship between INGO issue areas and their operational space.

Before defining our expectations, though, we must first note a few important caveats. As discussed previously, international NGOs have two avenues for legally working in China: (1) formal registration as an RO and (2) ad hoc temporary activity (TA) status. The RO process is more costly and requires far more administrative and legal work, and as a result, nearly twice as many organizations have used temporary activities rather than formal registration. In this study, however, we do not look at patterns in TA use to explore INGO operational space. There is an incredible amount of variation in how TAs are used. Some organizations submit dozens of TA requests, possibly as a type of workaround to obtaining RO status, while others use them as true temporary activities, submitting requests for one-off conferences or workshops. INGOs using TAs thus do not work well as a test case of the relationship between INGO operational space and the de facto implementation of the ONGO law. More permanent ROs are more appropriate. Additionally, we cannot observe or measure the activities of INGOs that have not attempted to establish operations in the country or who left the country following the implementation of the ONGO law in 2017. The population of INGOs working with full, permanent legal status is thus limited only to organizations that have already been vetted as "safe" and acceptable by the government. Even under these conditions, though, we should expect to see variation in operational space, due to heterogeneity in both (1) government preferences and INGO issues (i.e. not all registered INGOs work on the same issues, and not all issues are of equal value or importance to the regime), and (2) INGO preferences and resources (i.e. not all INGOs have the same goals, missions, or resources).

We use the number of provinces an INGO is authorized to work in as a proxy for organizational flexibility. While this is not a perfect measure---geographic scope is only one possible aspect of INGO operational space---it is an element of INGO flexibility that is directly influenced by China's ONGO law. Since INGOs are required to seek out separate PSUs for each province they are interested in working in, the count of authorized provinces partially reflects the alignment of INGO services with government preferences. Having a larger geographic scope provides concrete forms of INGO flexibility, like being able to more rapidly respond to changing needs in emergency situations such as the COVID-19 pandemic (e.g., a health-focused INGO can redirect staff to a province facing an outbreak), and allowing for a more diversified range of programs to assist with legal risk management when the definition of INGO activities is ambiguous (e.g., an INGO can pull back its work in a province if the relationship with its PSU sours). Previous work has shown that INGOs in China often work to maximize their geographic scope, even if they have no current plans to conduct activities nationwide. For instance, Oxfam and World Vision went through the cumbersome process to register multiple ROs to cover all the provinces they need to work in, and despite their goal of registering across mainland China, they faced increased bureaucratic hurdles and restriction with each registration [@Li:2020]. Based on @fig-theory, we posit that the alignment of government preferences and INGO services---represented by to contentiousness of an INGO's main issue area---helps determine INGO operational space. We thus expect that the number of provinces an organization is authorized to work in varies by INGO issue area:

> **E~1~**: Because issue areas reflect varied level of contentiousness and government preferences, the number of provinces INGOs are approved to operate in is associated with the contentiousness of their issue areas.

@@

@@

*add the discussion of "significance level, or the interpretation of credible intervals. That is move some of the notes to the figure to the main text?"* - explain credible intervals and how it shows us the whole distribution and is more actionable than confidence intervals and trying to get a single parameter. The point of this whole model is to paint a broad picture of the whole situation.
